---
title: "LINCS-inference"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{LINCS-inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In the previous vignettes, we've only looked at simulated data. In this demonstration, we're going to dive into some real dose-response data. These data was gathered as part of the NIH Library of Integrated Network-Based Cellular Signatures (LINCS), and are avaiable at https://lincs.hms.harvard.edu/.

The HMS LINCS data have been analyzed using metrics including the GR_{50} and GR_{max}, which are two metrics of growth rate inhibition proposed by Hafner et. al 2016. Both of these metrics depend on fitting a sigmoidal curve between the log of the drug concentration and the normalized cellular growth rate. 

One drawback of these methods is that they are unable to produce any quantification of uncertaintly. Therefore, it is possible for metrics like GR_{50} and GR_{max} to be driven by random noise, giving the appearance of a therapeutic effect when in reality there is none.

By using branching process inference with functional dependence on drug dose, we will provide a mechanistically-aware estimate of the effect of drug dose on cellular kinetics, along with robust uncertaintly quantifications.


First we load the `bpinference` package.

```{r setup}
library(bpinference)
```

Next, we'll define the structure of our model. This looks similar to `one_type_double_logistic.Rmd`. For now, we will only let the birth rate depend on drug concentration, and treat death rate as a fixed parameter. Feel free to re-run this model with both parameters varying as a function of dose. However, it may be difficult to effectively separate changes in birth and death due to the sparsity of the data.

$b = c_1 + \frac{c_2}{ 1 +\exp{(-c_3(x_1 - c_4))}}\\ d = c_5$

```{r}
e_mat <-  matrix(c(2,0),ncol=1)
p_vec <- c(1, 1)
func_deps = c('c[1] + c[2]/(1 + exp(c[3]*(x[1] - c[4])))','c[5]') # logistic birth functions
n_params <- 5
n_deps <- 1
mod = bp_model(e_mat, p_vec, func_deps, n_params, n_deps)
```

We now have a `bp_model` object which we can use for simulation and inference. Next we'll load in the LINCS data. The LINCS data file contains emprilical dose-response curves for 106 drugs and 6 different cell lines. You can download the data at https://lincs.hms.harvard.edu/db/.

```{r}
delete_cols = c("Timepoint.Unit","Small.Mol.Conc.Unit","LJP.Library.Plate","Assay.Well","Normalized.Growth.Rate.Inhibition.Value","Cell.HMS.LINCS.ID","Small.Molecule.HMS.LINCS.ID","Relative.Cell.Count")
lincs_data = read.csv("lincs-data/dataset_20245_20181218181013.csv")
lincs_data = lincs_data[,!names(lincs_data)%in%delete_cols]
```

We'll now write a function to extract one particular cell-drug combination for our analysis.  
```{r}
prepare_data <- function(cellname, drug){ 
  small_data = dplyr::filter(lincs_data, Cell.Name == cellname, Small.Molecule.Name == drug)
  controls = small_data$Total.Control.Cell.Count[1:max(small_data$Replicate)];
  drug_conc = log(c(rep(0.001, length(controls)), small_data$Small.Mol.Concentration))
  final_pop = c(controls, small_data$Total.Cell.Count.After.Treatment)
  times = c(rep( small_data$Timepoint[1], length(controls)), small_data$Timepoint)/24
  init_pop = c(rep(small_data$Total.Cell.Count.Before.Treatment[1], length(controls)), 
               small_data$Total.Cell.Count.Before.Treatment)
  new_data = data.frame(cbind(drug_conc = drug_conc, final_pop = final_pop, times = times, init_pop = init_pop))
  return(new_data)
}

cell_name = "MCF7"
drug = "Seliciclib"
small_data = prepare_data(cell_name, drug)
print(small_data)
```

\code{small_data} has 4 columns: The initial population of each replicate, the final population of each replicate, the drug concentration of each replicate, and the duration of each replicate. To convert this data into a format that our Stan model will accept, we will use the `create_stan_data` fuction:

```{r}
stan_dat = create_stan_data(mod, final_pop = matrix(small_data$final_pop), init_pop = matrix(small_data$init_pop), c_mat = matrix(small_data$drug_conc), times = small_data$times)
```

Now that we have our input data, we can define priors, generate a Stan model, and start sampling.

```{r}
priors = list()
priors[[1]] <- prior_dist(name="normal", params = c(0, .5), bounds = c(0,5))
priors[[2]] <- prior_dist(name="normal", params = c(gr_range, .5), bounds = c(0,5))
priors[[3]] <- prior_dist(name="normal",params=c(0,5), bounds=c(0,10))
priors[[4]] <- prior_dist(name="normal",params=c(0,5), bounds=c(-5,5))
priors[[5]] <- prior_dist(name="normal", params = c(0, .5), bounds = c(0,5))


generate(mod, priors, "lincs_birth_logistic.stan")

options(mc.cores = parallel::detectCores())

ranges <- matrix(rep(c(0,1), mod$nparams),ncol=2,byrow = T)
ranges[4,] <- c(-2,2)
ranges[3,] <- c(0,5)
init <- uniform_initialize(ranges, 4)


stan_mod <- rstan::stan_model(file = "lincs_birth_logistic.stan")
fit_data <- rstan::sampling(stan_mod, data = stan_dat, control = list(adapt_delta = 0.95), chains = 4, refresh = 1, init =init, iter=2000)

```

To visualize how well we did, we'll plot a cloud of growth rate (birth - death) curves that we sampled from the model, overlayed with empirical growth rates from the data:



