---
title: "LINCS-inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LINCS-inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bpinference)
```

In the previous vignettes, we've only looked at simulated data. In this demonstration, we're going to dive into some real dose-response data. These data was gathered as part of the NIH Library of Integrated Network-Based Cellular Signatures (LINCS), and are avaiable at https://lincs.hms.harvard.edu/.

The HMS LINCS data have been analyzed using metrics including the GR_{50} and GR_{max}, which are two metrics of growth rate inhibition proposed by Hafner et. al 2016. Both of these metrics depend on fitting a sigmoidal curve between the log of the drug concentration and the normalized cellular growth rate. 

One drawback of these methods is that they are unable to produce any quantification of uncertaintly. Therefore, it is possible for metrics like GR_{50} and GR_{max} to be driven by random noise, giving the appearance of a therapeutic effect when in reality there is none.

By using branching process inference with functional dependence on drug dose, we will provide a mechanistically-aware estimate of the effect of drug dose on cellular kinetics, along with robust uncertaintly quantifications.


First we load the `bpinference` package.

```{r setup}
library(ggplot2)
devtools::load_all()
library(rstan)
```

Next, we'll define the structure of our model. This looks similar to `one_type_double_logistic.Rmd`. For now, we will only let the birth rate depend on drug concentration, and treat death rate as a fixed parameter. Feel free to re-run this model with both parameters varying as a function of dose!

$b = c_1 + \frac{c_2}{ 1 +\exp{(-c_3(x_1 - c_4))}}\\ d = c_5$

```{r}
e_mat <-  matrix(c(2,0),ncol=1)
p_vec <- c(1, 1)
func_deps = c('c[1] + c[2]/(1 + exp(c[3]*(x[1] - c[4])))','c[5]') # logistic birth functions
n_params <- 5
n_deps <- 1
mod = bp_model(e_mat, p_vec, func_deps, n_params, n_deps)
```

We now have a `bp_model` object which we can use for simulation and inference. Next we'll load in the LINCS data. The LINCS data file contains emprilical dose-response curves for 106 drugs and 6 different cell lines.

```{r}
times = seq(0,5)
z0 = c(1000) # initial population vector
simulation_params = c(0.15, .2, 10, 0.6, 0.2, .12, 10, .4)
c_mat= matrix(c(0.0,0.2,0.4,0.6,0.8,1.0), ncol=1)
simulation_data = bpsims(mod, simulation_params, z0, times, rep(20,6), c_mat)
```

Next we define our priors, generate the Stan model, and prepare our sampling data.
```{r}
priors = rep(list(list(name="normal",params=c(0, .5), bounds=c(0,2))),8)
priors[[3]] = list(name="uniform",params=c(5,15), bounds=c(5,15))
priors[[7]] = list(name="uniform",params=c(5,15), bounds=c(5,15))
generate(mod, priors, "one_type_double_logistic.stan")

dat = stan_data_from_simulation(simulation_data, mod)
ranges = matrix(rep(c(0,1), length(simulation_params)),ncol=2,byrow = T)
ranges[3,] = c(5,15)
ranges[7,] = c(5,15)
init = uniform_initialize(ranges, 4)
```

Now all that's left is to do is sample. This model is a bit more complex than the simple birth-death model, so it will take significantly longer to finish sampling. Here's the code to run the sampler:

```{r}
options(mc.cores = parallel::detectCores())
stan_mod <- stan_model(file = "one_type_double_logistic.stan")
fit_data = sampling(stan_mod, data = dat, control = list(adapt_delta = 0.95), chains = 4, refresh = 1, init =init)
samples = data.frame(extract(fit_data))
```

To visualize how well we did, we'll plot a cloud of response curves that we sampled from the posterior and compare them with ground truth. 
```{r}
compute_birth_curve = 
  function(params, doses){sapply(doses, function(x){params[1] + params[2]/(1 + exp(params[3]*(x - params[4])))})}
compute_death_curve = 
  function(params, doses){sapply(doses, function(x){params[1] - params[2]/(1 + exp(params[3]*(x - params[4])))})}
sample_birth_curves = apply(samples[1:800,1:4], 1, compute_birth_curve, seq(0,1,length.out = 1000)) 
sample_death_curves = apply(samples[1:800,5:8], 1, compute_death_curve, seq(0,1,length.out = 1000)) 
bcurves = cbind(melt(data.frame(sample_birth_curves)),dose = seq(0,1,length.out = 1000))
dcurves = cbind(melt(data.frame(sample_death_curves)),dose = seq(0,1,length.out = 1000))

ggplot(dcurves, aes(x=dose, y=value, group = factor(variable))) + geom_line(aes(color="estimated death rate"),alpha=.03) + geom_line(data = bcurves,aes(color = "estimated birth rate"), alpha = .03) + ylab("parameter value") + geom_line(data = data.frame(cbind(dose = seq(0,1,length.out = 1000), value = compute_birth_curve(simulation_params[1:4],seq(0,1,length.out = 1000)),variable=1))) + geom_line(data = data.frame(cbind(dose = seq(0,1,length.out = 1000), value = compute_death_curve(simulation_params[5:8],seq(0,1,length.out = 1000)),variable=1))) + guides(colour = guide_legend(override.aes = list(alpha = 1)))
```


That looks pretty good! Notice how we've been able to tease apart changes in birth and death rates just from looking at total cell counts. Current widely-used methods in pharmacology can only measure changes in net growth rate, so using a Bayesian growth assay like this one could help give us a finer-grained understanding of how drugs act on various cell lines.

